\documentclass{article}

\def\npart {II}
\def\nyear {2017}
\def\nterm {Michaelmas}
\def\nlecturer{Dr H.\ Wilton}
\def\ncourse{Algebraic Topology}
\input{header}

\usetikzlibrary{knots}

% preamble
\newcommand{\id}{\mathrm{id}}
\numberwithin{nthm}{subsection}
% and here we go!

% hjrw2
% orw notes
% hatchers book
\begin{document}
\maketitle
\section{Introduction}
% trefoil picture goes here (trefoil ~> unknot)

% \begin{figure}[!h]
%     \centering
%     \begin{tikzpicture}
%         \begin{knot}[
%           clip width=6,
%           consider self intersections=true,
%           flip crossing=2,
%           ]
%           \strand[ultra thick, red] (0,2)   .. controls +(2.5,0)   and +(120:-2.5) ..
%                                     (210:2) .. controls +(120:2.5) and +(60:2.5)  ..
%                                     (-30:2) .. controls +(60:-2.5) and +(-2.5,0)  ..
%                                     (0,2);
%         \end{knot}
%     \end{tikzpicture}
%     \caption{Trefoil diagram}
% \end{figure}
\paragraph{Question 1} Is the trefoil really a knot?

How do we make this precise?
We can think of the trefoil and the unknot as continuous embeddings
\begin{equation*}
    f_i: S^1 \hookrightarrow \R^3
\end{equation*}
where $f_0$ corresponds to the trefoil, and $f_1$ corresponds to the unknot.

\paragraph{Question 1 (precise)} Is there a continuous map
\begin{equation*}
    F:S^1 \times [0, 1] \to \R^3
\end{equation*}
such that
\begin{enumerate}[(i)]
    \item $F(\theta, 0) = f_0(\theta)$, $F(\theta, 1) = f_1(\theta)$ $\forall \theta \in S^1$
    \item $F(\cdot, t_0): S^1 \to \R^3$ is injective, $\forall t_0 \in [0, 1]$
\end{enumerate}

We phrased unknotting as an \emph{extension problem}.  We can think about other, easier to state extension problems.
\paragraph{Question 2}
Consider $S^2 = \set{\vec{x} \in \R^3 | \norm{x} = 1}$, and $D^3 = \Set{\vec{x} \in \R^3 | \norm{x} \leq 1}$ along with the natural inclusion $i: S^2 \hookrightarrow D^3$.
Does there exist a continuous $f: D^3 \to S^2$ such that $f \circ i = \id_{S^2}$?
% picture if i can be bothered

This seems like a hard question!  So, let's instead consider this analogous question in algebra.
We take $S^2$ analogous to $\Z$, and $D^3$ analogous to $\{0\}$, the trivial group. From here, we get another question:
\paragraph{Question 3} Does there exist a group homomorphism $f: \{0\} \to \Z$ such that $\id_\Z = f \circ 0$?

This question seems much easier to solve! This is the essence of algebraic topology - we turn difficult questions in topology into easy questions about algebra.

\subsection{Examples and conventions}
Zoo of examples

\begin{itemize}
    \item Point $*$
    \item Circle $S^1$
    \item The circle generalises to the $n$-sphere, $S^{n-1} = \Set{x \in \R^n | \norm{x} = 1}$
    \item But it also can be used to produce the torus $T^2 = S^1 \times S^1$.
        % more damn pictures
    \item Which itself generalises into the genus-$g$ surface, $\Sigma_g$
    \item The torus can alternatively be found by identifying edges of a square, but by identifying them differently we can make a Klein bottle $K$
    \item Or the real projective plane $\mathbb{RP}^2$
\end{itemize}

In this course, the term map always refers to a continuous map.
To check that a map is continuous, we'll almost always use
\begin{lemma}[The Gluing Lemma]
    If $X = C_1 \cup C_2$, with $C_i$ closed, and $f: X \to Y$ is a function such that $f|_{C_1}$ and $f|_{C_2}$ are both continuous, then $f$ is continuous.
\end{lemma}

\begin{proof}
    Let $D \subseteq Y$ be closed.  Then
    \begin{equation*}
        f^{-1}(D) = (f|_{C_1})^{-1} (D) \cup (f|_{C_2})^{-1} (D)
    \end{equation*}
    is a finite union of closed sets, hence is closed.  Therefore, $f$ is continuous.
\end{proof}

\subsection{Cell complexes}
There are the spaces we can build by gluing.  The \emph{basic operation} we use is to attach an $n$-dimensional disc, as shown in the non-existent diagram.
Formally, if $f: S^{n-1} \to X$ is a continuous map, then we produce
\begin{equation}
    (X \sqcup D^n) / \sim
\end{equation}
where $\sim$ is an equivalence relation: $Y \sim f(y)$, for $y \in S^{n-1}$ and $f(y) \in X$.
\begin{defi}[Cell complexes]
    We define cell complexes by induction:
    \begin{enumerate}[(i)]
        \item A zero-dimensional cell complex is a finite discrete topological space
        \item An $n$-dimensional cell complex is constructed from an $(n-1)$ dimensional cell complex $X$ by attaching finitely many $n$-dimensional discs to $X$
    \end{enumerate}
\end{defi}

\section{Homotopy and the Fundamental Group}
\paragraph{Basic question} How can we prove that $X \ncong Y$?
The basic idea is to associate algebraic objects (for instance groups) to $X, Y$.
\clearpage
\subsection{Homotopy}
Throughout, we will denote $I = [0, 1] \subseteq \R$.

% def 2.1.1
\begin{ndef}\hypertarget{def:homotopy}
    Let $f, g: X \to Y$ be maps. A \textbf{homotopy} from $f$ to $g$ is a continuous map $H: X \times I \to Y$ such that $H(x, 0) = f(x)$ and $H(x, 1) = g(x) \; \forall x \in X$.
    We say that $f$ and $g$ are \textbf{homotopic}, and write $f \simeq g$ (or $f \simeq_H g$).
\end{ndef}

\begin{ndef}\hypertarget{def:homotopyRel}
    If $A \subseteq X$ and $\forall a \in A, \forall t \in I$, $H(a, t) = f(a) = g(a)$, then we say \emph{$f$ is homotopic to $g$ relative to $A$} and we write $f \simeq g \, \mathrm{rel} A$.
\end{ndef}

% prop 2.1.2
\begin{nprop}
    The relation $\simeq$  (rel $A$) is an equivalence relation.
\end{nprop}

\begin{proof}
    Reflexivity and symmetry are easy exercises, so they are omitted.  Transitivity: we have $f, g, h: X \to Y$, where $f \simeq_H g$ and $g \simeq_H' h$, and our goal is to build a homotopy $f \simeq h$.

    colourful pictures go here

    From the diagram, it's clear that the required homotopy is given by
    \begin{align*}
        H'' : X \times I &\to Y \\
        (x, t) &\mapsto
        \begin{cases}
            H(x, 2t) & 0 \leq t \leq \frac12 \\
            H'(x, 2t-1) & \frac12 \leq t \leq 1
        \end{cases}
    \end{align*}
    This is continuous from the gluing lemma.
\end{proof}

\begin{ndef}\hypertarget{def:homotopyEq}
    If we have $f: X \to Y$ and $g: Y \to X$ (diagram) with $g \circ f \simeq \id_X$ and $f \circ g \simeq \id_Y$, then we say that $X$ and $Y$ are \textbf{homotopy equivalent}, and $f, g$ are \textbf{homotopy equivalences}, and we write $X \simeq Y$.
\end{ndef}

\begin{eg}
    Take $X = \R^n$ and $Y = *$. Consider the trivial map $r: \R^n \to *$ and $i: * \to \R^n$, $i(*) = 0$. Then $r \circ i = \id_*$ and $i \circ r = 0: \R^n \to \R^n$
    Let $H: R^n \times I \to \R^n$ $(x, t) \mapsto tx$, so for $t=0$, H(., 0) = 0 and for $t=1$, $H(., 1) = \id_{\R^n}$.
\end{eg}

\begin{eg}
    This time, consider the spaces $\R^2 \setminus {0}$ and $S^1$, and the maps $r(x) = \frac{x}{\norm{x}}$  and inclusion $i$.  Then $r \circ i = \id_{S^1}$ and we can use the homotopy $H(x, t) = tx + (1-t) \frac{x}{\norm{x}}$ to show $i \circ r \simeq \id_{\R^2 \setminus {0}}$.  So, $\R^2 \setminus \{0\}$ and $S^1$ are homotopy equivalent.
\end{eg}

\begin{ndef}
    If $X$ is homotopy equivalent to a point, we say that $X$ is \textbf{contractible}.
\end{ndef}

% def 2.1.7
\begin{nlemma}
    If
    [diagram]
    are maps with $f_0 \simeq_H f_1$ and $g_0 \simeq_H'$ then $g_0 \circ f_0 \simeq g_1 \circ f_1$.
\end{nlemma}

\begin{proof}
    We show that both are homotopic to $g_0 \circ f_1$.  $g_0 \circ f_0 \simeq g_0 \circ f_1$ via $g_0 \circ H$ and $g_0 \circ f_1 \simeq g_1 \circ f_1$ via $H'(f_1(\cdot), \cdot)$.
\end{proof}

% prop 2.1.8
\begin{prop}
    Homotopy equivalence is an equivalence relation of spaces.
\end{prop}

\begin{proof}
    Symmetry and reflexivity are trivial. Transitivity [diagram]
    Suppose these are homotopy equivalences. We need to show $g \circ g' \circ f' \circ f \simeq \id_X$ and $f' \circ f \circ g \circ g' \simeq \id_Z$.
    Now,
    \begin{equation*}
        g \circ (g' \circ f') \circ f \simeq g \circ \id_Y \circ f \simeq g \circ f \simeq \id_X \\
    \end{equation*}
    and
    \begin{equation*}
        f' \circ (f \circ g) \circ g' \simeq f' \circ \id_Y \circ g' \simeq f' \circ g' \simeq \id_Z \\
    \end{equation*}
\end{proof}

% defi 2.1.9
\begin{ndef}[\hypertarget{def:defRet}{Deformation retraction}]
    If $i: A \hookrightarrow X$ is an inclusion, $r: X \to A$ is a continuous map such that $r \circ i = \id_A$ and $i \circ r \simeq \id_X$ rel $A$ then $A$ is called a \textbf{deformation retract} of $X$, and $r$ is a \textbf{deformation retraction}.
\end{ndef}

\begin{ndef}
    If $r: X \to A$ is a map such that $ r \circ i = \id_A$, then $A$ is a \textbf{retract} and $r$ is a retraction.
\end{ndef}

\begin{eg}
    If $X \neq \emptyset$, then for any $x \in X$, $r: X \to \{x\}$ is a retraction.
\end{eg}

Recall our basic goal was to prove that $X \not\simeq Y$ or that $X \not \cong Y$, and we will develop tools to do this.  For instance we can think about the spaces $X = *$ and $Y = * \sqcup *$.

% 2.2
\subsection{Paths}

% 2.2.1
\begin{ndef}
    A \textbf{path} in a space $X$ is a map $\gamma: I \to X$.
    A \textbf{path from $x_0$ to $x_1$} is such a map with $\gamma(0) = x_0$ and $\gamma(1) = x_1$, which we can denote by $\gamma: x_0 \snake x_1$.
    If $x_0 = x_1$, we say that $\gamma$ is a \textbf{loop}.
\end{ndef}

If $\gamma_0: x_0 \snake x_1$ and $\gamma_1: x_1 \snake x_2$ then the contraction of $\gamma_0$ and $\gamma_1$ is
\begin{equation*}
    (\gamma_0 \cdot \gamma_1) (t) =
    \begin{cases}
        \gamma_0(2t) & 0 \leq t \leq \frac12 \\
        \gamma_1(2t-1) & \frac12 \leq t \leq 1
    \end{cases}
\end{equation*}

If $\gamma: I \to X$ is a path, the inverse path $\gamma^{-1} = \bar{\gamma}$, given by $\bar{\gamma}(t) = \gamma(1-t)$.
For any $x_0 \in X$, the constant path
\begin{align*}
    C_{x_0}: I &\to X \\
             t &\mapsto x_0
\end{align*}

% defi 2.2.2
\begin{ndef}\hypertarget{def:pathComp}{Path components}
    If there is a path $\gamma:x_0 \snake x_1$, then we write $x_0 \sim x_1$.  By the above, this is an equivalence relation on $X$.  We write
    \begin{equation*}
        \pi_0(X) = X/\sim
    \end{equation*}
    and call the equivalence classes $[x]$ the \textbf{path components} of X.
    If there is only one path component, we say $X$ is path connected.
\end{ndef}

What's more, if $f: X \to Y$ is a map, we can define
\begin{align*}
    \pi_0(f): \pi_0(X) &\to \pi_0(Y)
    [x] &\mapsto [f(x)]
\end{align*}
It's easy to see that this is well-defined.

\begin{nprop}
    For [diagram] continuous maps.
    \begin{enumerate}[label=(\roman*)]
        \item if $f_0 \simeq f_1$, then $\pi_0(f_0) = \pi_0(f_1)$
        \item $\pi_0(f \circ f_0) = \pi_0(g) \circ \pi_0(f_0)$
        \item $\pi_0(\id_X) = \id_{\pi_0(X)}$
    \end{enumerate}
\end{nprop}

\begin{proof}
    \leavevmode
    \begin{enumerate}[label=(\roman*)]
        \item Let $f_0 \simeq_H f_1$. For any $x \in X$, $H(x, \cdot): I \to Y$ is a path from $f_0(x) to f_1(x)$, hence $[f_0(x)] = [f_1(x)]$, and therefore $\pi_0(f_0)([x]) = \pi_0(f_1)([x])$.
    \end{enumerate}
    (ii) and (iii) are immediate.
\end{proof}

\begin{eg}[Toy example]
    We try to show $* \not \simeq * \sqcup *$.  First, see that $1 = \abs{\pi_0(*)}$ and $\abs{\pi_0(* \sqcup *)} = 2$. Suppose we have a pair of homotopy equivalences between $X$ and $Y$ called $f, g$.  Then $\pi_0(f) \circ \pi_0(g) = \pi_0(f \circ g) = \pi_0(\id_*) = \id_{\pi_0(*)}$. Likewise, $\pi_0(g) \circ \pi_0(f) = \Id_{\pi_0(* \sqcup *)}$ hence $\pi_0(f)$ is a bijection, $\implies 1=2$.
\end{eg}

Our next goal is to fix up the group axioms for paths. We could allow unrestricted homotopies of paths, but then all paths would be homotopic. Instead we fix the endpoints.

% defi 2.2.7
Paths $\gamma, \gamma': I \to X$ are \textbf{homotopic as paths} if $\gamma \simeq \gamma'$ rel $\{0, 1\}$.

Homotopy of paths and concatenations interact nicely.

% lemma 2.2.8
\begin{nlemma}
    If $\gamma_0, \gamma_0': x_0 \snake x_1$, (vert) $\gamma_1, \gamma_1': x_1 \snake x_2$ and $\gamma_0 \simeq \gamma_0'$, $\gamma_1 \simeq \gamma_1'$ as paths then $\gamma_0 \cdot \gamma_1 \simeq \gamma_0' \cdot \gamma_1'$.
\end{nlemma}

\begin{proof}[Sketch]
    Suppose $\gamma_0 \simeq_{H_0} \gamma_0'$ and $\gamma_1 \simeq_{H_1} \gamma_1'$.
    The diagram shows the homotopy of paths we are looking for.
\end{proof}

% prop 2.2.9
\begin{nprop}
    Let $\gamma_0:x_0 \snake x_1$, $\gamma_1: x_1 \snake x_2$, $\gamma_2: x_2 \snake x_3$.
    Then
    \begin{enumerate}[label=(\roman*)]
        \item $(\gamma_0 \cdot \gamma_1) \cdot \gamma_2 \simeq \gamma_0 \cdot (\gamma_1 \cdot \gamma_2)$
        \item $\gamma_0 \cdot c_{x_1} \simeq \gamma_0 \simeq c_{x_0} \cdot \gamma_0$
        \item $\gamma_0 \cdot \overline{\gamma_0} \simeq c_{x_0}
    \end{enumerate}
\end{nprop}
\end{document}
